% Glossary entries for the Summa commentary
% Please put in alphabetical order!.
\newglossaryentry{analogical}{
  name={Analogical},
  description={
    A comparison between two things. See analogy.
  }
}

\newglossaryentry{essence}{
  name={Essence},
  description={
    The fundamental, unchanging properties that define what a thing is, its core nature, independent of physical existence or human perception, contrasting with temporary or accidental attributes. The what-it-is-ness of a     thing that makes it a member of a category.
  }
}

\newglossaryentry{habit}{
  name={Habit},
  description={
    An effect from repeated act producing an aptitude to reproduce them.}=
  }
}


\newglossaryentry{name}{
  name={Name},
  description={
    A term used to signify the essence or nature of a thing.
    In metaphysics, to name something is to attempt to signify
    what it is. When angels reveal their names, this indicates
    their office or function: Raphael signifies God's healing,
    and Michael signifies the question “Who is like God?”
    God, however, cannot be named according to His essence,
    since His essence exceeds the comprehension of the created
    intellect. Nevertheless, we apply names to God analogically,
    according to the perfections found in creatures.
  }
}

\newglossaryentry{similitude}{
  name={Similitude},
  description={
    The quality of being similar to something else.
  }
}

\newglossaryentry{univocally}{
  name={Univocally},
  description={
    To express something in a clear unambiguous manner. Much unlike the use of univocally in common parlance.
  }
}
